\begin{thebibliography}{99}
\addcontentsline{toc}{chapter}{Bibliography}
	\bibitem{Box94} Box, G. E. P., Jenkins G. M. and Reinsel, G. C. (1994) Time Series Analysis, Forecasting and Control, 3rd edn., Eaglewood Cliffs: Prentice Hall.
	
	\bibitem{Dagum06} Dagum, E. B. and Cholette, P. A. (2006) Benchmarking, Temporal Distribution, and Reconciliation Methods for Time Series, \textit{Lecture Notes in Statistics}, Volume \textbf{186}, page 15, Springer New York.
	
	\bibitem{DJIA} YAHOO! Finance, Dow Jones Industrial Average -- Historical prices,\\
	\url{http://finance.yahoo.com/q/hp?s=^DJI}.
	
	\bibitem{dofactory-Decorator} data \& object factory, \textit{Decorator},\\
	\url{http://www.dofactory.com/Patterns/PatternDecorator.aspx}.
	
	\bibitem{Faraway97} Faraway J. and Chatfield C. (1997) Time Series Forecasting with Neural Networks: A Comparative Study Using the Airline Data, \textit{Applied Statistics}, Volume \textbf{47}, Issue 2 (1998), 231-250.
	
	\bibitem{Kaastra95} Kaastra, I. and Boyd, M. (1995) Designing a neural network for forecasting financial and economic time series, \textit{Neurocomputing}, Volume \textbf{10} (1996) 215-236.
	
	\bibitem{Klimasauskas93} Klimasauskas, C. C. (1993) Applying neural networks, in R. R. Trippi and E. Turban, eds., \textit{Neural Networks in Finance and Investing: Using Artificial Intelligence to Improve Real World Performace} (Probus, Chicago) 64-65.
	
	\bibitem{Masters93} Masters, T. (1993) Practical Neural Network Recipes in C++, Academic Press, New York.
	
	\bibitem{Socha04} Socha, K. (2004) ACO for Continuous and Mixed-variable Optimization, \textit{Lecture Notes in Computer Science}, Volume \textbf{3172}, 25--36, Springer Berlin / Heidelberg.
	
	\bibitem{Sima96} Šíma J. a Neruda R. (1996) \textit{Teoretické otázky neuronových sítí}, Matfyzpress, Praha.
	
	\bibitem{Tang91} Tang Z., de Almeida C. and Fishwick, P. A. (1991) Time series forecasting using neural networks versus Box-Jenkins methodology, \textit{Simulation}, \textbf{57}, 303-310.
	
	\bibitem{Vasicek} Vašíèek, Z. Simulované žíhání (Simmulated Annealing), Metropolis\r{u}v algoritmus.
	
	\bibitem{Whitley94} Whitley, D. (1994) A Genetic Algorithm Tutorial, \textit{Statistics and Computing}, Volume \textbf{4}, 65--85.
	
	\bibitem{Wikipedia-Artificial_neural_network} Wikipedia -- The Free Encyclopedia, \textit{Artificial neural network},\\
	\url{http://en.wikipedia.org/wiki/Artificial_neural_network}.
	
	\bibitem{Wikipedia-Feedforward_neural_network} Wikipedia -- The Free Encyclopedia, \textit{Feedforward neural network}\\
	\url{http://en.wikipedia.org/wiki/Feedforward_neural_network}.
	
	\bibitem{Wikipedia-Stigmergy} Wikipedia -- The Free Encyclopedia, \textit{Stigmergy},\\
	\url{http://en.wikipedia.org/wiki/Stigmergy}.
	
	\bibitem{Zhang04} Zhang, G. P. et al. (2004) Neural Networks in Business Forecasting, Idea Group Publishing, London.
\end{thebibliography}